---
title: "Project title"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
```

# Data 1

## Introduction and data

-   Identify the source of the data.

For this project proposal, we will be exploring County Demographics. The source of this dataset is from the CORGIS Dataset Project, specifically by Ryan Whitcomb, Joung Min Choi, and Bo Guan.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

This dataset was originally collected from 2010 to 2019 by the United States Census Bureau, which individually surveys all residents of the United States.

-   Write a brief description of the observations.

The observations describe the "age distributions, the education levels, employment statistics, ethnicity percents, household information, income, and other miscellaneous statistics" of all U.S. counties from 2010 to 2019.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-1

# add code here
```

# Data 2

## Introduction and data

-   Identify the source of the data.

For this project we will be exploring English Premier League Football data. The source of this data comes from Organization called Football.CSV which explores football data from all over the world.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

The data was collected by recording every game that every team in the Premier League played over the past 30 years.

-   Write a brief description of the observations.

The observation describe "Date, Round number, FT score, Team 1, Team 2, Standing Position", for every game played in the Premier League from 2010-2019.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-2

flights <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-07-12/flights.csv')

```

# Data 3

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-3

# add code here
```
