---
title: "Project Proposal - Team 199"
subtitle: "Amalia Desir, Luke Francese, Ella Davis, Ruben Mesalles"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
```

# Data 1

## Introduction and data

-   Identify the source of the data.

For this project proposal, we will be exploring County Demographics. The source of this dataset is from the CORGIS Dataset Project, specifically by Ryan Whitcomb, Joung Min Choi, and Bo Guan.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

This dataset was originally collected from 2010 to 2019 by the United States Census Bureau, which individually surveys all residents of the United States.

-   Write a brief description of the observations.

The observations describe the "age distributions, the education levels, employment statistics, ethnicity percents, household information, income, and other miscellaneous statistics" of all U.S. counties from 2010 to 2019.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-1

county_demographics <- readr::read_csv('https://think.cs.vt.edu/corgis/datasets/csv/county_demographics/county_demographics.csv')

# add glimpse. the glimpse will not run with the current code 

# add code here
```

# Data 2

## Introduction and data

-   Identify the source of the data.

For this project we will be exploring English Premier League Football data. The source of this data comes from Organization called Football.CSV which explores football data from all over the world.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

The data was collected by recording every game that every team in the Premier League played over the past 30 years.

-   Write a brief description of the observations.

The observation describe "Date, Round number, FT score, Team 1, Team 2, Standing Position", for every game played in the Premier League from 2010-2019.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-2

flights <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-07-12/flights.csv')

```

# Data 3

## Introduction and data

-   The data is a cleaned dataset created by a user on Kaggle (https://www.kaggle.com/datasets/heesoo37/120-years-of-olympic-history-athletes-and-results)

-   The original source of the data is http://www.sports-reference.com/, which has a compilation of data from the Olympic games over the years.

-   The dataset has observations for each athlete competing in an event, and it has variables such as team, age, sex, and sport.

## Research question

-   Which Olympic skiing event's results can best be predicted by athletes' characteristics, such as nationality, sex, weight, and height?

-   This research topic will explore which Olympic skiing victories are the most predictable based on the characteristics of competitors. This question could be important in knowing which events are the most predictable, and which may be more contested, which could guide coverage of the events. We predict that the events that tend to be dominated by one country, such as freestyle skiing (dominated by Canada) will have the best predictive models.

-   The question involves categorical variables such as event, country, sex, and medal, as well as quantitative variables such as height and weight.

## Literature

-   https://pietistschoolman.com/2014/02/21/which-country-dominates-which-olympic-sport-winter-edition/

-   The article shows the most dominated winter Olympic sports, broken down both by country and continent. This provides a decent basis to form a hypothesis, as particularly dominated sports may be more easily predicted than tossup sports. The article also suggests that continent, and not just country, could be a basis to sort athletes in order to predict results.

-   The article does not go into depth regarding whether any of the events can be predicted year-to-year, so it can easily be expanded on.

## Glimpse of data

```{r}
#| label: load-data-3

olympics <- read.csv("data/olympics.csv") 
glimpse(olympics)
```
