---
title: "Project title"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)

```

# Data 1

## Introduction and data

-   Identify the source of the data.

For this project proposal, we will be exploring County Demographics. The source of this dataset is from the CORGIS Dataset Project, specifically by Ryan Whitcomb, Joung Min Choi, and Bo Guan.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

This dataset was originally collected from 2010 to 2019 by the United States Census Bureau, which individually surveys all residents of the United States.

-   Write a brief description of the observations.

The observations describe the "age distributions, the education levels, employment statistics, ethnicity percents, household information, income, and other miscellaneous statistics" of all U.S. counties from 2010 to 2019. The county_demographics dataset includes rows by county across America and columns such as "Age.Percent 65 and Older", "Age.Percent Under 18 Years", "Education.High School or Higher", and "Ethnicities.Two or More Races".

## Research question

-   A well-formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

Do factors---such as race, age, or education level---impact the median household incomes of counties across America?

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

The research topic will not only explore whether or not the factors stated above influence counties' median household income, but it will also dive into equity and inequity as it relates to economic disparity across the country. We will research related topics, such as income inequality and rural-urban divide, to get a bigger picture of what the data truly means in the broader United States landscape.

Our hypothesis is that counties with higher ages and education levels will, in turn, produce higher median household incomes. We do believe that factors such as age, race, and education level influence median household incomes.

-   Identify the types of variables in your research question. Categorical? Quantitative?

Variables in this dataset that are categorical include, but are not limited to, "Education.High School or Higher", "Ethnicities.White Alone", and "Ethnicities.Two or More Races". Variables in this dataset that are quantitative include, but are not limited to, "Age.Percent 65 and Older", "Population.Population per Square Mile", and "Age.Percent Under 18 Years".

## Literature

-   Find one published credible article on the topic you are interested in researching.

One published credible article we found was: [Understanding Equity Through Census Bureau Data](https://www.census.gov/library/stories/2021/09/understanding-equity-through-census-bureau-data.html)

-   Provide a one paragraph summary about the article.

This article reviews the US Census Bureau Data by looking at several different reports that point to overall societal inequality. It details the Income and Poverty report, the Supplemental Poverty Measure, female-to-male earning ratios, health insurance coverage, Annual Education Attainment, household wealth, home ownership, and Community Resilience Estimates. These overviews help readers, analysts, and researchers understand what impact the pandemic has made on topics in the United States that connect to equity.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

Our research question expands beyond the article cited because it does not solely look at data pre- and post-pandemic. The data we will analyze relies on pre-pandemic (2010-2019) statistics that will help us gauge the state of counties 4 years ago and lead us to draw our own similarities and comparisons with what we know currently. Overall, we want to gain a much broader sense of overall trends, using a 10-year dataset to conduct research, instead of focusing on a smaller portion of recent history.

## Glimpse of data

```{r}
#| label: load-data-1

county_demographics <- readr::read_csv('https://think.cs.vt.edu/corgis/datasets/csv/county_demographics/county_demographics.csv')
```

# Data 2

## Introduction and data

-   Identify the source of the data.

The source of this data comes from FIFA World Cup.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

This data set covers every single World Cup match played in its history from Uruguay in 1930 to Russia in 2018.

-   Write a brief description of the observations.

> The wcmatches dataset contains every football match played in the World Cup. Columns inside the dataset include `city` from which city was the match being `played` in, `outcome` which team `won`, `drew`, or `loss`. Also shows if the winning side need added `extra time of penalties` to win the game. The data set also includes the `date`, `stage` and more.
>
> Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

    Do certain factors, such as the number of days between matches or the location of the match, impact the likelihood of a team winning a match during the FIFA World Cup tournament, and do these factors vary depending on the round of the tournament?

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

The research topic for this study is to investigate the factors that may influence a team's success in the `FIFA World Cup tournament`. We are interested in examining the `impact` of the `number of days` between matches and the `location` of the match on a team's likelihood of `winning` a match, and whether these factors `diffe`r depending on the `round` of the tournament.

Our `hypothesis` is that teams who have `more days between matches` may have an `advantage`, as they have more time to `rest` and recover from previous matches. Additionally, we hypothesize that the `location` of the match may also have an `impact` on a team's success. We also believe that the `impact` of these factors may `vary` depending on the `round` of the tournament, as teams may have different levels of `fatigue` or `pressure` depending on how far they have `progressed` in the tournament.

-   Identify the types of variables in your research question. Categorical? Quantitative?

    The number of days between matches: quantitative. The location of the match: categorical. The round of the tournament: categorical

## Literature

-   Find one published credible article on the topic you are interested in researching.

Weston M, Castagna C, Impellizzeri FM, Rampinini E, Abt G. Analysis of physical match performance in English Premier League soccer referees with particular reference to first half and player work rates. J Sci Med Sport. 2007 Dec;10(6):390-7. doi: 10.1016/j.jsams.2006.09.001. Epub 2006 Nov 28. PMID: 17126077.

-   Provide a one paragraph summary about the article.

The article investigates the impact of `player work rate` on `match outcome` in professional English soccer. The study uses data from `20 Premier League matches` and analyzes player movement data to evalaute y player `work rates`. The results indicate that `higher levels of work rate` are associated with a greater likelihood of a `positive match outcome`. The study also finds that the impact of player work rate on match outcome `may differ` depending on the `position played` by the player. We can understand how different factors, such as `work rate` and `position`, impact a team's performance, which can be used `improve` their chances of success.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

While this article does not focus specifically on the `FIFA World Cup tournament`, it provides evidence into the `impact` of player `work rate` on match outcomes in professional soccer, which may be `relevant` to the research question I proposed earlier.

## Glimpse of data

```{r}
#| label: load-data-2

wcmatches <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-11-29/wcmatches.csv')
worldcups <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-11-29/worldcups.csv')

```

# Data 3

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-3

# add code here
```
