---
title: "Team 199: Olympics"
subtitle: "Report"
format: html
editor: visual
execute:
  echo: false
---

# Introduction

Throughout our project, we want to explore the relationship between Olympic athletes' traits and successful performance in swimming. We aim to define which traits---particularly nationality, BMI, weight, and height---influence athletes' performance. We seek to answer the question: which trait ultimately affects Olympic swimmers the most? Within this question, we investigate event, country, sex, medal, height, and weight. We hypothesize that height and nationality will impact Olympic swimmers the most. Filtering the data to address Olympic swim meets from the year 2000 and beyond, we aim to answer our research question via visually comprehensible plots and logistic models.

# Research question & Data

Which traits can best predict Olympic swimmers' performance (between nationality, height, weight, age, and BMI)?

The data is a cleaned dataset created by a user on Kaggle (https://www.kaggle.com/datasets/heesoo37/120-years-of-olympic-history-athletes-and-results), and it originated from http://www.sports-reference.com/, which has a compilation of data from the Olympic games over the years. The dataset has observations for each athlete competing, and it has variables such as team, age, sex, and sport. For each column, the brief definitions are as follows:

1.  **ID** - Unique number for each athlete

2.  **Name** - Athlete's name

3.  **Sex** - M or F

4.  **Age** - Integer

5.  **Height** - In centimeters

6.  **Weight** - In kilograms

7.  **Team** - Team name

8.  **NOC** - National Olympic Committee 3-letter code

9.  **Games** - Year and season

10. **Year** - Integer

11. **Season** - Summer or Winter

12. **City** - Host city

13. **Sport** - Sport

14. **Event** - Event

15. **Medal** - Gold, Silver, Bronze, or NA

# Literature

One published credible article on the topic that we are interested in researching is an article published by Swimming World Magazine, a US-based monthly swimming magazine that was first published in January 1960. It concurrently runs online websites Swimming World Magazine and Swimming World News, and the headquarters is in Phoenix, Arizona. The particular article that applies to our research is their March 2023 article, "What Makes the Perfect Swimmer's Body?".

Since our research is seeking the physical traits that have the strongest correlation with successful swimmer performance, this article acts as a foundation for our inspiration and our hypothesis.

In this article, the author JP Mortenson, an alumnus of Dartmouth's swimming and diving team and a physical therapist for professional swimmers, lists traits that many professional and Olympic swimmers have that have aided in their past athletic successes.

One of these traits is height, which is also found in our data set. JP claims hat height tends to give swimmers a distinct advantage in the pool, as made apparent by elite international male and female swimmers such as Nathan Adrian (6'6), Kilment Kolesnikov(6'5), Michael Andrew (6'5), Sarah Sjostrom (6'0), Olympic gold medalist Simone Manuel (5'11), and Olympic gold medalist Katie Ledeckey (5'11). The author even goes on to state that the average height of the 2016 Olympic finalists in Rio was 6'2 for men and 5'9 for women, which is 5 inches above the global average male and female heights respectively (worlddata.info and ourworldindata.org). Other traits that seem to lead to athletic success in the sector of swimming that were listed in this article include anthropometry, center of mass, flexibility, hyperextension abilities, and mental strength.

Our research question, "Which trait ultimately affects Olympic swimmers' performance the most?" builds upon this article because it would lend more concrete statistical evidence to many observed correlations such as height. Although this article provides evidence to its claims by citing corresponding examples of athletes, our research question will be able to build upon these claims by calling more physical traits into question, and specific swimming events in which these traits have the strongest correlations with success.

Article: https://www.swimmingworldmagazine.com/news/what-makes-the-perfect-swimmers-body/

# Methodology

The differences between swimmers are best shown through data visualization. The data was filtered to show only results from the 1992 Summer games and the years since, to make it more applicable to modern games while maintaining a large sample size. The relationship between national origin and medal results was shown through a stacked bar plot that shows the number of each type of medals earned by each country with at least ten medals.

```{r gearing_up_data}

library(tidyverse)
library(tidymodels)
library(lubridate)

olympics <- read.csv("data/olympics.csv") 

olympics_new <- olympics |> 
  mutate(BMI = (Weight/(Height/100)^2), medalyesno = if_else(is.na(Medal), "no", "yes"))

olympics_new <- as.data.frame(olympics_new)
```

```{r country_comparisons}

olympics_new2 <- olympics_new|> 
  mutate(Team = if_else(Team == "Unified Team", "Russia", Team))

medal_colors <- c("Gold" = "#FFD700", "Silver" = "#C0C0C0", "Bronze" = "#CD7F32")

olympics_new2 |> 
  mutate(Team = if_else(Team == "Unified Team", "Russia", Team)) |> 
  filter(Sport == "Swimming", Year > 1991, medalyesno == "yes") |> 
  group_by(Team) |> 
  filter(n() > 9) |> 
  count(Team, Medal, name = "Frequency") |> 
  mutate(Medal = factor(Medal, levels = c("Gold", "Silver", "Bronze"))) |> 
  group_by(Team) |> 
  mutate(Total_Medals = sum(Frequency)) |> 
  ungroup() |> 
  ggplot(aes(x = reorder(Team, -Total_Medals), y = Frequency, fill = Medal)) +
  geom_col() +
  scale_fill_manual(values = medal_colors) +
  labs(x = "Team", y = "Number of Medals", fill = "Medal", 
       title = "Olympic Swimming Medals by Country", subtitle = "countries with 10+ medals, 1992-2016") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

These visualizations show a very strong relationship between national origin and Olympic swimming performance. Australian and American swimmers had by far the majority of all medal types, while the majority of countries had a negligible number. National origin is thus a strong predictive factor for swimming performance.

Next, parallel boxplots were used to compare the ranges for athlete's physical traits and age, comparing different medalists to non-medalists. Each plot was faceted by age to determine if trends in age or physical characteristics differed for males and females.

```{r athlete_stats}

olympics_new2 |> 
  filter(Sport == "Swimming", Year > 1991) |>
  mutate(medal = as.factor(Medal), medal = fct_relevel(Medal, "Gold", "Silver", "Bronze")) |>
  ggplot(aes(x = BMI, y = medal)) + 
  geom_boxplot() + 
  facet_wrap(~Sex) + 
  labs(title = "BMI vs. Medal Earned", 
       subtitle = "for Women and Men, 2000-2016 Olympics",
       x = "BMI of Athlete", 
       y = "Type of Medal Earned")

olympics_new2 |> 
  filter(Sport == "Swimming", Year > 1991) |>
  mutate(medal = as.factor(Medal), medal = fct_relevel(Medal, "Gold", "Silver", "Bronze")) |>
  ggplot(aes(x = Height, y = medal)) + 
  geom_boxplot() + 
  facet_wrap(~Sex) + 
  labs(title = "Height vs. Medal Earned", 
       subtitle = "for Women and Men, 2000-2016 Olympics",
       x = "Height of Athlete (cm)", 
       y = "Type of Medal Earned")

olympics_new2 |> 
  filter(Sport == "Swimming", Year > 1991) |>
  mutate(medal = as.factor(Medal), medal = fct_relevel(Medal, "Gold", "Silver", "Bronze")) |>
  ggplot(aes(x = Weight, y = medal)) + 
  geom_boxplot() + 
  facet_wrap(~Sex) + 
  labs(title = "Weight vs. Medal Earned", 
       subtitle = "for Women and Men, 2000-2016 Olympics",
       x = "Weight of Athlete (Kg)", 
       y = "Type of Medal Earned")

olympics_new2 |> 
  filter(Sport == "Swimming", Year > 1991) |>
  mutate(medal = as.factor(Medal), medal = fct_relevel(Medal, "Gold", "Silver", "Bronze")) |>
  ggplot(aes(x = Age, y = medal)) + 
  geom_boxplot() + 
  facet_wrap(~Sex) + 
  labs(title = "Age vs. Medal Earned, 2000-2016 Olympics", 
       subtitle = "for Women and Men",
       x = "Age of Athlete", 
       y = "Type of Medal Earned")

```

The typical BMI, height, and weight for swimmers increased slightly as their medal rank did (so the typical gold winner was slightly taller and heavier, and had a higher BMI than the other medalists as well as non-medalists). These results are conserved across sex in each category. The age of the swimmer had no meaningful correlation with results; there are slight differences between medal categories but no large effect or clear trend.

A logistic regression model was chosen to determine which variables have the strongest effect on swimmer performance. For this model, all medals were merged into one category, and the model attempted to predict whether or not an athlete would win any medal based on their traits. Based on the visualization results, Age was excluded as a variable to be tested, leaving Team, Height, Weight, and BMI. The model only included the countries that had earned ten or more medals, making the assumption that athletes from other countries had a negligible likelihood of winning.

```{r}

olympics3 <- olympics_new2 |> 
  mutate(Team = if_else(Team == "Unified Team", "Russia", Team)) |> 
  filter(Sport == "Swimming",
         Year > 1991,
         Team == c("United States", "Australia", 
                   "Japan", "Netherlands", 
                   "France", "China", 
                   "Germany", "Great Britain", 
                   "Russia", "Canada", 
                   "Italy", "Hungary", 
                   "South Africa", "Sweden")) |> 
  mutate(medalyesno = as.factor(medalyesno)) |> 
  drop_na(Height)


logistic_swim <- logistic_reg() |>
    set_engine("glm") |>
    fit(medalyesno ~ Team,
        data = olympics3, family = "binomial")

glance(logistic_swim)

logistic_swim2 <- logistic_reg() |>
    set_engine("glm") |>
    fit(medalyesno ~ Team + BMI,
        data = olympics3, family = "binomial") 

glance(logistic_swim2) #this made AIC worse 

logistic_swim3 <- logistic_reg() |>
    set_engine("glm") |>
    fit(medalyesno ~ Team + Weight, 
        data = olympics3, family = "binomial")

glance(logistic_swim3) #slightly better 

logistic_swim4 <- logistic_reg() |>
    set_engine("glm") |>
    fit(medalyesno ~ Team + Weight + Height, 
        data = olympics3, family = "binomial") 

glance(logistic_swim4) #best yet

logistic_swim5 <- logistic_reg() |>
    set_engine("glm") |>
    fit(medalyesno ~ Height + Weight, 
        data = olympics3, family = "binomial")

glance(logistic_swim5) #testing without Team. Just awful

logistic_swim6 <- logistic_reg() |>
    set_engine("glm") |>
    fit(medalyesno ~ Team + Height, 
        data = olympics3, family = "binomial")

glance(logistic_swim6) #testing without Weight. Better than with. 

logistic_swim7 <- logistic_reg() |> 
    set_engine("glm") |>
    fit(medalyesno ~ Height, 
        data = olympics3, family = "binomial")

glance(logistic_swim7) #testing with just Height. much worse than just Team 
```

This set of logistic models yielded two models with comparable AIC values: Team + Height + Weight and Team + Height. Models that did not incorporate the Team variable had far higher AIC than those that did, while the model with Team alone had an AIC close to the final model. This makes it clear that Team is the variable that best predicts medaling. The p-values of the two best models were checked to determine which yielded more significant results.

```{r final_model}

tidy(logistic_swim6) |> 
  knitr::kable() 

tidy(logistic_swim4) |> 
  knitr::kable() 

```

```{r}

logistic_swim90 <- logistic_reg() |>
    set_engine("glm") |>
    fit(medalyesno ~ Height + Team,  
        data = olympics3, family = "binomial")
glance(logistic_swim90)
tidy(logistic_swim90) |> 
  knitr::kable() 

logistic_swim8 <- logistic_reg() |>
    set_engine("glm") |>
    fit(medalyesno ~ Height,  
        data = olympics3, family = "binomial")
glance(logistic_swim8)
tidy(logistic_swim8) |> 
  knitr::kable() 

logistic_swim9 <- logistic_reg() |>
    set_engine("glm") |>
    fit(medalyesno ~ Team,  
        data = olympics3, family = "binomial")
glance(logistic_swim9)
tidy(logistic_swim9) |> 
  knitr::kable() 
```

$$\log\Big(\frac{p}{1-p}\Big) = -7.5364541 + 0.0401009 \times Height + [TeamCoef.]
$$

$$p = \frac{\exp(-7.5364541 + 0.0401009\times Height\ + [TeamCoef.])}{1 + \exp(-7.5364541 + 0.0401009\times Height\ + [TeamCoef.])}$$

# Results

All of the tested characteristics except for age showed a significant effect on medal results, with a clear trend between placement and each characteristic. High-placing swimmers tended to be taller, heavier, and of a higher BMI than low-placing swimmers. The effect of the physical characteristics, however, was far overshadowed by the effect of national origin. The typical ranges for gold-medal swimmers had significant overlap with those of non-medalists in most comparisons, while dominant countries such as the United States and Australia claimed the vast majority of medals and had a far greater fraction of winners as well.

These results support the idea that national support and the training regimen for Olympic swimmers is far more important than the specifics of their physical traits. Because the medal data favored certain countries, it could be proposed that the main factors that support an Olympian's success are country-specific factors such as the popularity of their sport or the availability of funding for them to train. While physical characteristics are still important (the data showed that only people with a narrow range of body measurements were even in the competition), national support and other factors are critical for high-level success.
