---
title: "Team 199: Olympics"
subtitle: "Report"
format: html
editor: visual
execute:
  echo: true
---

# Introduction

Throughout our project, we want to explore the correlation between athletes' traits and performance in swimming. We aim to define which traits---particularly nationality, sex, weight, and height---influence athletes' performance. We seek to answer the question: which trait ultimately affects Olympic swimmers the most? This question involves categorical variables such as event, country, sex, and medal, as well as quantitative variables such as height and weight. We hypothesize that height and nationality will impact Olympic swimmers the most.

# Research question & Data

Which traits can best predict Olympic swimmers' performance (between nationality, height, weight, age, and BMI)?

The data is a cleaned dataset created by a user on Kaggle (https://www.kaggle.com/datasets/heesoo37/120-years-of-olympic-history-athletes-and-results), and it originated from http://www.sports-reference.com/, which has a compilation of data from the Olympic games over the years. The dataset has observations for each athlete competing, and it has variables such as team, age, sex, and sport. For each column, the brief definitions are as follows:

1.  **ID** - Unique number for each athlete

2.  **Name** - Athlete's name

3.  **Sex** - M or F

4.  **Age** - Integer

5.  **Height** - In centimeters

6.  **Weight** - In kilograms

7.  **Team** - Team name

8.  **NOC** - National Olympic Committee 3-letter code

9.  **Games** - Year and season

10. **Year** - Integer

11. **Season** - Summer or Winter

12. **City** - Host city

13. **Sport** - Sport

14. **Event** - Event

15. **Medal** - Gold, Silver, Bronze, or NA

# Literature

      One published credible article on the topic that we are interested in researching is an article published by Swimming World Magazine, a US-based monthly swimming magazine that was first published in January 1960. It concurrently runs online websites Swimming World Magazine and Swimming World News, and the headquarters is in Phoenix, Arizona. The particular article that applies to our research is their March 2023 article, "What Makes the Perfect Swimmer’s Body?". 
      Since our research is seeking the physical traits that have the strongest correlation with successful swimmer performance, this article acts as a foundation for our inspiration and our hypothesis. 
      In this article, the author JP Mortenson, an alumnus of Dartmouth's swimming and diving team and a physical therapist for professional swimmers, lists traits that many professional and Olympic swimmers have that have aided in their past athletic successes. 
      One of these traits is height, which is also found in our data set. JP claims hat height tends to give swimmers a distinct advantage in the pool, as made apparent by elite international male and female swimmers such as Nathan Adrian (6'6), Kilment Kolesnikov(6'5), Michael Andrew (6'5), Sarah Sjostrom (6'0), Olympic gold medalist Simone Manuel (5'11), and Olympic gold medalist Katie Ledeckey (5'11). The author even goes on to state that the average height of the 2016 Olympic finalists in Rio was 6'2 for men and 5'9 for women, which is 5 inches above the global average male and female heights respectively (worlddata.info and ourworldindata.org). Other traits that seem to lead to athletic success in the sector of swimming that were listed in this article include anthropometry, center of mass, flexibility, hyperextension abilities, and mental strength. 
      Our research question, "Which trait ultimately affects Olympic swimmers' performance the most?" builds upon this article because it would lend more concrete statistical evidence to many observed correlations such as height. Although this article provides evidence to its claims by citing corresponding examples of athletes, our research question will be able to build upon these claims by calling more physical traits into question, and specific swimming events in which these traits have the strongest correlations with success. 
      Article: https://www.swimmingworldmagazine.com/news/what-makes-the-perfect-swimmers-body/
      

# Methodology

The differences between swimmers are best shown through data visualization. The data was filtered to show only results from the 2000 Summer games and the years since, to make it more applicable to modern games. The relationship between national origin and medal results was shown in two ways: a first plot shows how many swimming medals of each kind had gone to each country, while the second plot shows the proportion of competitors from the top seven countries

```{r gearing_up_data}

library(tidyverse)
library(tidymodels)
library(lubridate)

olympics <- read.csv("data/olympics.csv") 

olympics_new <- olympics |> 
  mutate(BMI = (Weight/(Height/100)^2), medalyesno = if_else(is.na(Medal), "no", "yes"))

olympics_new <- as.data.frame(olympics_new)
```

```{r country_comparisons}

olympics_new |> 
  filter(Sport == "Swimming",
         medalyesno == "yes",
         Year > 1999) |>
  mutate(medal = as.factor(Medal),
         medal = fct_relevel(Medal, "Gold", 
                             "Silver", "Bronze")) |>
  ggplot(aes(x = Team, fill = medal)) + 
    geom_bar() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

olympics_new |> 
  filter(Sport == "Swimming",
         Year > 1999,
         Team == c("Australia", "China", "France",
                   "Germany", "Japan", "Netherlands",
                   "United States")) |>
  mutate(medal = as.factor(Medal), 
         medal = if_else(str_detect(medal, "NA"), "No Medal", "olympics_new"),
         medal = fct_relevel(Medal, "Gold", "Silver", "Bronze", "No Medal")) |>
  ggplot(aes(x = Team, fill = medal)) + 
    geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  facet_wrap(~Sex) 
```

```{r}

olympics_new |> 
  filter(Sport == "Swimming", Year > 1999) |>
  mutate(medal = as.factor(Medal), medal = fct_relevel(Medal, "Gold", "Silver", "Bronze")) |>
  ggplot(aes(x = BMI, y = medal)) + 
    geom_boxplot() + 
  facet_wrap(~Sex) + 
  labs(title = "BMI vs. Medal Earned", 
       subtitle = "for Women and Men",
       x = "BMI of Athlete", 
       y = "Type of Medal Earned")

olympics_new |> 
  filter(Sport == "Swimming", Year > 1999) |>
  mutate(medal = as.factor(Medal), medal = fct_relevel(Medal, "Gold", "Silver", "Bronze")) |>
  ggplot(aes(x = Height, y = medal)) + 
    geom_boxplot() + 
  facet_wrap(~Sex) + 
  labs(title = "Height vs. Medal Earned", 
       subtitle = "for Women and Men",
       x = "Height of Athlete (cm)", 
       y = "Type of Medal Earned")

olympics_new |> 
  filter(Sport == "Swimming", Year > 1999) |>
  mutate(medal = as.factor(Medal), medal = fct_relevel(Medal, "Gold", "Silver", "Bronze")) |>
  ggplot(aes(x = Weight, y = medal)) + 
    geom_boxplot() + 
  facet_wrap(~Sex) + 
  labs(title = "Weight vs. Medal Earned", 
       subtitle = "for Women and Men",
       x = "Weight of Athlete (Kg)", 
       y = "Type of Medal Earned")

olympics_new |> 
  filter(Sport == "Swimming", Year > 1999) |>
  mutate(medal = as.factor(Medal), medal = fct_relevel(Medal, "Gold", "Silver", "Bronze")) |>
  ggplot(aes(x = Age, y = medal)) + 
    geom_boxplot() + 
  facet_wrap(~Sex) + 
  labs(title = "Age vs. Medal Earned", 
       subtitle = "for Women and Men",
       x = "Age of Athlete", 
       y = "Type of Medal Earned")

```

# Results
